\textls[-15]{\thetitle \mbox{(Technical Report)}}

### 2022-06-26 coqimpl.tex
took a pass on partial-recursor paras

\textbf{Propositional Partial Recursor.}
Apparently, the exposed module "b_3" in \cref{fig:plugin-example2} is
not expressive enough to really fulfill \ref{langdesign-req2}. For
example, we cannot prove the constructors are injective, i.e.,
\mintinline{Coq}{self_B.tt ≠ self_B.ff}. In the original setting, to
prove injectivity of constructor of (vanilla) inductive type, we only
require a recursor and its computational rules. However, as we pointed
out earlier, the recursor of an inductive type cannot be exposed into
the context for the sake of extensibility. Thus we need to alter the
formulation of the recursor i.e., we need to have a better
characterization of ``extensible inductive type'' exported in the
compiled module type as a part of the interface for the following fields
to use.

The simplest idea is to allow partiality---i.e. returning \mintinline{Coq}{option T} instead of the original \mintinline{Coq}{T}. We call this \textit{propositional partial recursor}. For example, the propositional partial recursor for \mintinline{Coq}{B.b} is 
\begin{minted}{Coq}
b_prec : forall P : b -> Set, P tt -> P ff 
    -> forall x : b, option (P x). 
\end{minted}
with appropriate computational axioms. The partiality doesn't break extensibility (i.e. all the future extension of "b" can support this "b_prec").

What's more, partial recursor is theoretically enough for the "injection" and "discriminate" tactics. The rough idea to achieve this is to use "b_prec" to reflect our special inductive type into vanilla inductive type, and use the orignal "injection" and "discrimination" respectively. For example, we want to prove \mintinline{Coq}{self_B.tt ≠ self_B.ff}, we simply 
\begin{minted}{Coq}
Definition reflect : self_B.b → option bool 
                   := b_prec (fun _ => bool) true false.
Definition tt_neq_ff : self_B.tt = self_B.ff → False.
(* Because self_B.tt = self_B.ff → reflect self_B.tt = reflect self_B.ff 
                → true = false    (by computational axiom)
                → False           (by discriminate of bool) *)
\end{minted}

In fact, the above "reflect" function implies that propositional partial recursor is good enough to some extent, at least for non-indexed inductive type, because we can ``embed'' vanilla inductive type into types supporting partial recursor. Because "reflect" is actually a left inverse of an injection "bool → self_B.b". A bit more formally, and restrict our focus to non-indexed inductive types,

\begin{theorem}\label{thm:prec-complete} Given a list of $n$ pairs of types $\{ x : "A"_i \vdash "B"_i(x) \}_{i}$
  \begin{itemize}
    \item Define $"C"$ as the inductive type using this list, i.e. 
    \begin{minted}{Coq}
Inductive C : Set := ... | cᵢ : forall (x : Aᵢ), (Bᵢ x -> C) -> C | ...
    \end{minted}
    \item Assume an arbitrary type \mintinline{Coq}{D : Set} that has ...
    \begin{itemize}
      \item following $n$ \textbf{functions} \mintinline{Coq}{dᵢ : forall (x : Aᵢ), (Bᵢ x -> D) -> D}  
      \item a partial recursor 
      \begin{minted}{Coq}
prec : forall (R : Set), ..,  
  (rᵢ : (forall (x : Aᵢ), (Bᵢ x -> R) -> R)), .., D -> R
      \end{minted}
      \item all the following $n$ (propositional) computational axioms: for all $i$, 
      \begin{minted}{Coq}
prec {R} r₁ r₂ .. rₙ (dᵢ aᵢ bᵢ) 
    = rᵢ aᵢ (fun x => prec {R} r₁ r₂ .. rₙ (bᵢ x))
      \end{minted}
    \end{itemize}
  \end{itemize}
  We will have an \textbf{embedding} from "C" to "D", more concretely, we can have \mintinline{Coq}{inj : C -> D} and

  \mintinline{Coq}{linv : D -> option C} defined using eliminator and "prec", (both act similar to identity) such that
  \begin{minted}{Coq}
    forall c : C, linv (inj c) = some c
  \end{minted}
\end{theorem}

We emphasize that in \cref{thm:prec-complete}, each function $"d"_i$ can actually be considered as a constructor, because they can be reflected to real constructors of "C" using "linv". Thus, \cref{thm:prec-complete} implies that (1) every future extension of the inductive type can support this partial recursion (trivially); (2) every type support this partial recursor with its computational axiom at least support these constructors because of the embedding. In other words, \textit{a type (at least) supports these constructors if and only if this type supports the corresponding partial recursor (with the computational axiom)}.  Thus we can argue that the partial recursor gives a sound and complete characterization of all the extension of a given (non-indexed) inductive type.

>>>>

\textbf{Propositional Partial Recursor.}
Apparently, the exposed module "b_3" in \cref{fig:plugin-example2} is
not expressive enough to really fulfill \ref{langdesign-req2}. For
example, we cannot prove the constructors are injective, i.e.,
\mintinline{Coq}{self_B.tt ≠ self_B.ff}. In the original setting, to
prove injectivity of constructors of a (vanilla) inductive type, we only
require a recursor and its computational rules. However, as we pointed
out earlier, the recursor of an inductive type cannot be exposed into
the context for the sake of extensibility. Hence, we need to alter the
formulation of the recursor: we need to have a better
characterization of ``extensible inductive types'' exported in the
compiled module type as a part of the interface for the following fields
to use.

The simplest idea is to allow partiality---i.e., returning
\mintinline{Coq}{option T} instead of the original \mintinline{Coq}{T}.
We call this \textit{propositional partial recursor}.
For example, the propositional partial recursor for \mintinline{Coq}{B.b} is 
\begin{minted}[escapeinside=@@]{Coq}
b_prec : forall P : b@\YZ{b or self_B.b?}@ -> Set, P tt -> P ff
    -> forall x : b, option (P x).
\end{minted}
with appropriate computational axioms. The partiality doesn't break
extensibility (i.e. all future extensions of "b" can support this
"b_prec").

What's more, partial recursor is theoretically enough for the
"injection" and "discriminate" tactics. The rough idea to achieve this
is to use "b_prec" to reflect our special inductive type into vanilla
inductive type, and use the original "injection" and "discriminate"
tactics. For example, we want to prove \mintinline{Coq}{self_B.tt ≠
self_B.ff}, we simply 
\begin{minted}{Coq}
Definition reflect : self_B.b → option bool 
                   := b_prec (fun _ => bool) true false.
Definition tt_neq_ff : self_B.tt = self_B.ff → False.
(* Because self_B.tt = self_B.ff → reflect self_B.tt = reflect self_B.ff 
                → true = false    (by computational axiom)
                → False           (by discriminate of bool) *)
\end{minted}

In fact, the above "reflect" function implies that propositional partial
recursor is good enough to some extent, at least for non-indexed
inductive type, because we can ``embed'' vanilla inductive types into
types supporting a partial recursor. Because "reflect" is actually a left
inverse of an injection "bool → self_B.b". A bit more formally, and
restricting our focus to non-indexed inductive types,

\begin{theorem}\label{thm:prec-complete} Given a list of $n$ pairs of types $\{ x : "A"_i \vdash "B"_i(x) \}_{i}$
  \begin{itemize}
    \item Define $"C"$ as the inductive type using this list, i.e.,
    \begin{minted}{Coq}
Inductive C : Set := ... | cᵢ : forall (x : Aᵢ), (Bᵢ x -> C) -> C | ...
    \end{minted}
    \item Assume an arbitrary type \mintinline{Coq}{D : Set} with
    \begin{itemize}
      \item $n$ \textbf{functions} \mintinline{Coq}{dᵢ : forall (x : Aᵢ), (Bᵢ x -> D) -> D}  
      \item a partial recursor
      \begin{minted}[escapeinside=@@]{Coq}
prec : forall (R : Set), ..,  @\YZ{Can R be indexed?}@
  (rᵢ : (forall (x : Aᵢ), (Bᵢ x -> R) -> R)), .., D -> R
      \end{minted}
      \item and $n$ (propositional) computational axioms: for all $i$, 
      \begin{minted}{Coq}
prec {R} r₁ r₂ .. rₙ (dᵢ aᵢ bᵢ) 
    = rᵢ aᵢ (fun x => prec {R} r₁ r₂ .. rₙ (bᵢ x))
      \end{minted}
    \end{itemize}
  \end{itemize}
  Then there exists an \textbf{embedding} from "C" to "D".
  More concretely, we can have \mintinline{Coq}{inj : C -> D} and 
  \mintinline{Coq}{linv : D -> option C} defined using eliminator and "prec"
  (both acting like identity) such that
  \begin{minted}{Coq}
    forall c : C, linv (inj c) = some c
  \end{minted}
\end{theorem}

We emphasize that in \cref{thm:prec-complete}, each function $"d"_i$ can
actually be considered as a constructor, because they can be reflected
to real constructors of "C" using "linv". Thus, \cref{thm:prec-complete}
implies that (1) every future extension of the inductive type can
support this partial recursion (trivially); (2) every type supporting
this partial recursor with its computational axioms at least supports
these constructors because of the embedding. In other words, \textit{a
type (at least) supports these constructors if and only if this type
supports the corresponding partial recursor (with the computational
axioms)}.  Thus, we can argue that the partial recursor gives a sound and
complete characterization of all the extension of a given (non-indexed)
inductive type.\YZ{What about indexed inductive types?}
